Computing {\em the least commitment} action model from state observations

### 2018-11-27 doc/ijcai-19-conformant/conformant-learning.tex
added definition for the explanation of an observation, ijcai19 paper

\section{Introduction}
Given a sequence of partially observed states, this paper formalizes the task of computing the exact set of action models that are {\em conformant} with the given observation. This task is of interest because it allows to, incrementally and scalably, learn action models from arbitrary large sets of state observations.

In addition, the paper presets a new method to compute {\em the least commitment} action model from state observations. The method assumes that action models are specified as \strips\ action schema and builds on top of off-the-shelf {\em conformant planning} algorithms.

\section{Background}
This section formalizes the {\em classical} and {\em conformant} planning models as well as the kind of input observations for the computation of {\em the least commitment} action model.

\subsection{Classical planning with conditional effects}
$F$ is the set of {\em fluents} or {\em state variables} (propositional variables). A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. $L$ is a set of literals that represents a partial assignment of values to fluents, and $\mathcal{L}(F)$ is the set of all literals sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is a full assignment of values to fluents. We explicitly include negative literals $\neg f$ in states s.t. $|s|=|F|$ and the size of the state space is $2^{|F|}$.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state in which all the fluents of $F$ are assigned a {\em true\textbar false} value and $G\in\mathcal{L}(F)$ is the set of goal conditions over the state variables. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, and $|\pi|=n$ denotes its {\em plan length}. The execution of $\pi$ in the initial state $I$ of $P$ induces a {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ solves $P$ iff the induced {\em trajectory} $\tau(\pi,s_0)$ reaches a final state $G \subseteq s_n$.

@@

A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,s_0)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, range from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.

We assume a bijective monotone mapping between actions/states of trajectories and observations~\cite{ramirez2009plan}, thus also granting the inverse consistency relationship (the trajectory is a superset of the observation). Therefore, transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

\subsection{Conformant planning}
{\em Conformant planning} refers to planning with incomplete information about the initial state, no sensing, and where goals have to be achieved with certainty (despite the uncertainty of the initial state)~\cite{smith1998conformant,goldman1996expressive}.

Syntactically, conformant planning problems are expressed in compact form through a set of state variables. A {\em conformant planning problem} is then defined as a tuple $P_c=\tup{F,A,\Upsilon,G}$ where $F$, $A$ and $G$ are the set of fluents, actions and goals (as previously defined for the classical planing model). Now $\Upsilon$ is a set of clauses over literals $l=f$ or $l=\neg f$ (for $f\in F$) that define the set of possible initial states.

A solution to a conformant planning problem is an action sequence that maps each possible initial state into a goal state. More precisely, an action sequence $\pi=\tup{a_1, \ldots, a_n}$ is a {\em conformant plan} for $P_c$ iff for each possible {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$, such that $s_0$ is a valuation of the fluents in $F$ that satisfies $\Upsilon$, then the $\tau(\pi,s_0)$ trajectory reaches a final state $G \subseteq s_n$.

\section{Learning {\em the least commitment} action model from state observations}
Now we are ready to formalize the learning of the {\em the least commitment} action model from state observations and our approach to address it via {\em conformant planning}.

\subsection{The {\em least commitment} action model}
The task of computing {\em the least commitment} action model is defined as a tuple $\tup{P,M,\mathcal{O}}$ where:

\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G[\cdot]}$ is a planning problem where $A[\cdot]$ is a set of actions. For each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined. The set of goal conditions $G[\cdot]$ is also unknown.
\item $M$ is the set of different planning models for the actions in $A[\cdot]$. A model $\mathcal{M}\in M$ defines the semantics of every action in $A[\cdot]$. Planning models differ in the $\tup{\rho,\theta}$ functions of the actions but they all use the same set of state variables $F$.
\item $\mathcal{O}(\tau)$ is a sequence of state observations coming from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$.
\end{itemize}

The {\em solution} to this task is {\em the least commitment} action model which defines the smallest subset of models $M^*\subseteq M$ that is {\em conformant} with the input observation $\mathcal{O}(\tau)$.

\subsection{The space of possible action models}
This work focuses on the lerning of action models that are specified as \strips\ action schema. \strips\ action schemata provide a compact representation for specifying action models. {\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

\begin{definition}[Comparable \strips\ action schemata]
Two \strips\ schemata $\xi$ and $\xi'$ are {\bf comparable} iff $pars(\xi)=pars(\xi')$, i.e, both share the same list of parameters.\footnote{In \strips\ models, $pars(\xi)=pars(\xi')$ implies the number of parameters must be the same. For other planning models that allow object typing, the equality implies that parameters share the same type}
\end{definition}

For instance, the {\small\tt stack(?v1,?v2)} and {\small\tt unstack(?v1,?v2)} schemata from a four operator {\em blocksworld}~\cite{slaney2001blocks} are {\em comparable} while {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function that maps every action schema $\xi\in\mathcal{M}$ to a comparable schemata $\xi'\in\mathcal{M'}$ and vice versa.

@@

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because of the semantic of the {\tt\small on} predicate (i.e. only one block can be on top of another block). Invariants constraining the space of possible states in a given domain belong also to this group~\cite{fox1998automatic}. 
\end{enumerate}

>>>>

\section{Introduction}
Given a sequence of partially observed states, this paper formalizes the task of computing {\em the least commitment} action model that can explain the given observation. This task is of interest because it allows the incremental learning of action models from arbitrary large sets of state observations.

In addition, the paper introduces a new method to compute {\em the least commitment} action model from a sequence of state observations. The method assumes that action models are specified as \strips\ action schema and it builds on top of off-the-shelf algorithms for {\em conformant planning}.

\section{Background}
This section formalizes the model of {\em classical planning with conditional effects}, the {\em conformant planning} model, and the kind of input observations for the computation of {\em the least commitment} action model.

\subsection{Classical planning with conditional effects}
Let $F$ be the set of {\em fluents} or {\em state variables} (propositional variables). A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. $L$ is a set of literals that represents a partial assignment of values to fluents, and $\mathcal{L}(F)$ is the set of all literals sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is a full assignment of values to fluents. We explicitly include negative literals $\neg f$ in states s.t. $|s|=|F|$ and the size of the state space is $2^{|F|}$.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state and $G\in\mathcal{L}(F)$ is the set of goal conditions over the state variables. A {\em plan} $\pi$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, with $|\pi|=n$ denoting its {\em plan length}. The execution of $\pi$ in the initial state $I$ of $P$ induces a {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ solves $P$ iff the induced {\em trajectory} $\tau(\pi,s_0)$ reaches a final state $G \subseteq s_n$. A solution plan is optimal iff its length is minimal.

@@

A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,s_0)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, ranges from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.

We assume a bijective monotone mapping between trajectories and observations~\cite{ramirez2009plan}, thus also granting the inverse consistency relationship (the trajectory is a superset of the observation). Therefore, transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

\subsection{Conformant planning}
{\em Conformant planning} is planning with incomplete information about the initial state, no sensing, and where goals have to be achieved with certainty (despite this uncertainty of the initial state)~\cite{smith1998conformant,goldman1996expressive}.

Syntactically, conformant planning problems are expressed in compact form through a set of state variables. A {\em conformant planning problem} can be defined as a tuple $P_c=\tup{F,A,\Upsilon,G}$ where $F$, $A$ and $G$ are the set of {\em fluents}, {\em actions} and {\em goals} (as previously defined for {\em classical planning}). Now $\Upsilon$ is a set of clauses over literals $l=f$ or $l=\neg f$ (for $f\in F$) that define the set of possible initial states.

A solution to a conformant planning problem is an action sequence that maps each possible initial state into a goal state. More precisely, an action sequence $\pi=\tup{a_1, \ldots, a_n}$ is a {\em conformant plan} for $P_c$ iff for each possible {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$, such that $s_0$ is a valuation of the fluents in $F$ that satisfies $\Upsilon$, then the $\tau(\pi,s_0)$ trajectory reaches a final state $G \subseteq s_n$ in which all the goal conditions are met.

\section{Learning {\em the least commitment} action model from state observations}
This section formalizes the notion of {\em the least commitment action model} that is able to {\em explain} a sequence of state observations and our approach to compute such model via {\em conformant planning}.

\subsection{The {\em least commitment} action model}
The task of computing {\em the least commitment} action model is defined as a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a planning problem where $A[\cdot]$ is a set of actions. For each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined. 
\item $M$ is the set of possible models for the actions in $A[\cdot]$. A model $\mathcal{M}\in M$ defines the semantics of every action in $A[\cdot]$. Planning models $\mathcal{M}\in M$ differ in the $\tup{\rho,\theta}$ functions of the actions but they all use the same set of state variables $F$.
\item $\mathcal{O}(\tau)$ is a sequence of partial states coming from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$ that solves $P$.
\end{itemize}

The {\em solution} to this task is {\em the least commitment} action model but, before formalizing it, we introduce the following necessary definition.

\begin{definition}[Explanation of an observation]
Given a model $\mathcal{M}\in M$ for the actions $A[\cdot]$ in $P$, and a sequence of partially observed states $\mathcal{O}(\tau)$, we say that {\em the model explains the observation}, that we denote by $\mathcal{M}\mapsto\mathcal{O}(\tau)$, iff there exists a plan $\pi$ that solves $P$ and that it is consistent with $\pi$. We say that $\pi$ is the {\em best explanation} for $\mathcal{O}(\tau)$ iff, in additon, $\pi$ is optimal.
\end{definition}

Now we are ready to define {\em the least commitment} action model as the largest subset of models $M^*\subseteq M$ such that every model $\mathcal{M}\in M^*$ explains the input observation $\mathcal{O}(\tau)$.

\subsection{The space of possible action models}
In spite of previous definitions are general, this work focuses on action models that are specified as \strips\ action schemata. \strips\ action schemata provide a compact representation for specifying action models. {\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

@@

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because of the semantic of the {\tt\small on} predicate (i.e. only one block can be on top of another block). As a rule of thumb, {\it invariants} constraining the possible states of a given planning domain belong to this class of constraints~\cite{fox1998automatic}. 
\end{enumerate}
